Constructing a universe for the setoid model

### 2020-01-08 types20-abstract.tex
Add missing citations

The setoid interpretation gives rise to a model of a type theory with functional
and propositional extensionality. It is thus a way to explain extensionality in
a type-theoretic and computationally adequate way \cite{alti99}. The setoid
model has been recently adapted into Setoid Type Theory (SeTT), which is
justified by a syntactic translation into a very basic \emph{target theory}.
\cite{mpc19}. This translation relies on the existence in the target theory of a
universe of definitionally proof-irrelevant propositions $\Prop$, as recently
implemented in Coq and Agda \cite{Sprop}. The setoid interpretation models a
universe of propositions where equality of propositions is logical equivalence,
thus providing a very basic instance of Univalence. However, we also would like
to be able to equip SeTT with a universe of setoids. This universe must in turn
be a setoid, hence in particular it cannot be univalent, but rather its
equivalence relation must reflect simple equality of codes.

@@

Note that this construction gives rise to an inductive-recursive universe in
$\Set_1$, rather than $\Set$, thus the meta-theory must contain at least one
universe. Our result is that a modified form of this translation also works for
the more complex inductive-recursive type we need to model the universe of
setoids. 
%
In particular, in addition to $\ad{in-U}$ for defining $\U$ as before, we also
introduce a family $\ad{in-U∼}$ of binary relations between types in the
universe, from which we then define $\_\ad{∼U}\_$.
%
However, we now need an inductive-inductive type in the target theory.
%
\begin{align*}
  & \textsf{data}\ \ad{in-U} : \Set \to \Set_1 \\
  & \textsf{data}\ \ad{in-U∼} : \{A\ A' : \Set\} \to \ad{in-U}\ A \to \ad{in-U}\ A' \to (A \to A' \to \Prop) \to \Set_1 \\
  & \setoidU : \Set_1 \\
  & \El : \setoidU \to \Set \\
  & \_\sim_{\setoidU}\_ : \setoidU \to \setoidU \to \Prop_1 \\
  & \_\vdash\_\sim_{\El}\_ : \{a\ a' : \U\} \to a\ \sim_{\setoidU}\ a' \to \El\ a \to \El\ a' \to \Prop \\
  & \\
  & \U = \Sigma\ (X : \Set)\ (\AgdaDatatype{in-U}\ X) \\
  & \El = \textsf{proj}_1 \\
  & (X , p) \sim_{\setoidU} (X' , p') =
   \ \parallel \Sigma\ (R : X \to X' \to \Prop) (\ad{in-U~} p\ p'\ R) \parallel
\end{align*}

where $\parallel A \parallel\ : \Prop$ is propositional truncation for any $A :
\Set$. Intuition would suggest to define $\_\vdash\_\sim_{\El}\_$ by projecting
the relation out of the proof of $a \sim_{\setoidU} a'$, in much the same way as
$\El$ is defined by projecting out of $\U$. That is, $p \vdash x \sim_{\El} y =
\ad{proj}_1\ p\ x\ y$.
%
However, this doesn't work since the type of $p$ is propositionally truncated,
hence it cannot be used to construct a proof-relevant object. We can work around
this by defining $\_\vdash\_\sim_{\El}\_$ explicitly for all type formers, by
induction on the codes $a\ a' : \U$, mutually with a proof that this definition
is in some sense equivalent to what we would get if we could project out of the
truncation.

@@

\bibliography{bibliography}

>>>>

The setoid interpretation gives rise to a model of a type theory with functional
and propositional extensionality. It is thus a way to explain extensionality in
a type-theoretic and computationally adequate way \cite{setoid-model}. The
setoid model has been recently adapted into Setoid Type Theory (SeTT), which is
justified by a syntactic translation into a very basic \emph{target theory}
\cite{mpc19}. This translation relies on the existence in the target theory of a
universe of definitionally proof-irrelevant propositions $\Prop$, as recently
implemented in Coq and Agda \cite{gilbert}. The setoid interpretation models a
universe of propositions where equality of propositions is logical equivalence,
thus providing a very basic instance of Univalence. However, we also would like
to be able to equip SeTT with a universe of setoids. This universe must in turn
be a setoid, hence in particular it cannot be univalent, but rather its
equivalence relation must reflect simple equality of codes.

@@

Note that this construction gives rise to an inductive-recursive universe in
$\Set_1$, rather than $\Set$, thus the meta-theory must contain at least one
universe. Our result is that a modified form of this translation also works for
the more complex inductive-recursive type we need to model the universe of
setoids. 
%
In particular, in addition to $\ad{in-U}$ for defining $\U$ as before, we also
introduce a family $\ad{in-U∼}$ of binary relations between types in the
universe, from which we then define $\_\sim_{\setoidU}\_$.
%
However, we now need an inductive-inductive type in the target theory.
%
\begin{align*}
  & \textsf{data}\ \ad{in-U} : \Set \to \Set_1 \\
  & \textsf{data}\ \ad{in-U∼} : \{A\ A' : \Set\} \to \ad{in-U}\ A \to \ad{in-U}\ A' \to (A \to A' \to \Prop) \to \Set_1 \\
  & \setoidU : \Set_1 \\
  & \El : \setoidU \to \Set \\
  & \_\sim_{\setoidU}\_ : \setoidU \to \setoidU \to \Prop_1 \\
  & \_\vdash\_\sim_{\El}\_ : \{a\ a' : \U\} \to a\ \sim_{\setoidU}\ a' \to \El\ a \to \El\ a' \to \Prop \\
  & \\
  & \setoidU = \Sigma\ (X : \Set)\ (\AgdaDatatype{in-U}\ X) \\
  & \El = \textsf{proj}_1 \\
  & (X , p) \sim_{\setoidU} (X' , p') =
   \ \parallel \Sigma\ (R : X \to X' \to \Prop) (\ad{in-U∼}\ p\ p'\ R) \parallel
\end{align*}

where $\parallel A \parallel\ : \Prop$ is propositional truncation for any $A :
\Set$. Intuition would suggest to define $\_\vdash\_\sim_{\El}\_$ by projecting
the relation out of the proof of $a \sim_{\setoidU} a'$, in much the same way as
$\El$ is defined by projecting out of $\U$. That is, $p \vdash x \sim_{\El} y =
\ad{proj}_1\ p\ x\ y$.
%
However, this doesn't work since the type of $p$ is propositionally truncated,
hence it cannot be used to construct a proof-relevant object. We can work around
this by defining $\_\vdash\_\sim_{\El}\_$ explicitly for all type formers, by
induction on the codes $a\ a' : \U$, mutually with a proof that this definition
is logically equivalent to what we would obtain if we could project out of the
truncation.

@@

\bibliographystyle{alpha}
\bibliography{bibliography}
